Evaluating isolation and contact tracing for control of influenza outbreaks with pandemic potential: a modelling study

### 2025-06-19 inst/paper/main.tex
Changes from 19 June meeting

Outbreaks of severe novel infectious diseases pose a major public health threat. The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. Recent human cases of avian influenza A/H5N1, alongside transmission, illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece]. It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.

@@

The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host. If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time. If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}. Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}. If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off. In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\ % AK: Note 'primary case' means first in outbreak

Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease. These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g. Sharma et al, Nature Comms]. The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with a short generation times (e.g. pandemic influenza), incidence of new infections can grow quickly, outpacing a response. Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}. \\

@@

Several pre-COVID-19 studies  concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]

In this study we evaluate the effectiveness of controlling infectious disease outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model. In effect the ability to control an epidemic with isolation and contact tracing is a race between the identification and isolation of contacts and the spread of the infectious agent \citep{reyna-laraVirusSpreadContact2021}. The dogma that influenza contagion outpaces our ability to trace and isolate in time, may not hold for all pathogen subtypes with new estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}. This study looks to define the speed and effectiveness of isolation under different response delays across epidemic scenarios to inform outbreak response; quantifying the impact of reducing the delay to isolation on outbreak containment and the utility of rapid testing.

@@

To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}). The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility). Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission. The mean of the negative binomial offspring distribution is the basic reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}). When a case is symptomatic they are isolated after a delay. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). Any infection times that are sampled to occur after the isolated of an infector are remove, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times. In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$). Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated. The isolation of the infectee is whatever occurs first of these two events.  \\

@@

@@

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The contol of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

@@

>>>>

Outbreaks of severe respiratory infections pose a major public health threat. The COVID-19 pandemic resulted in over 7 million reported deaths \citep{whocovid-19dashboardCOVID19DeathsWHO}, while the 1918 influenza pandemic caused an estimated $\sim$50 million deaths \citep{johnsonUpdatingAccountsGlobal2002}. 
Influenza subtypes to which populations have low levels of pre-existing immunity are a particular concern as potential pandemic candidates. This includes influenza A/H5N1 \citep{Ward2024.12.11.24318702}, A/H7N9 \citep{tannerPandemicPotentialAvian2015}, and A/H1N1p when it caused caused a pandemic in 2009 \citep{fraserPandemicPotentialStrain2009}. Recent human cases of avian influenza A/H5N1, alongside transmission, further illustrate the ongoing potential for viruses to infect new hosts, creating the potential for further adaption [CITE: Antia et al, Nature, 2003 + key papers on H5N1 - see my PLOS Bio piece]. It is therefore crucial to understand the effectiveness and feasibility of different intervention strategies in the face of such infections.

@@

%The likelihood of a novel outbreak causing sustained transmission depends both the characteristics of the pathogen and its host population, as well control measures  implemented in response to early cases. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases by a typical infectious host. If the reproduction number is greater than 1, we would expect the incidence of infections to increase over time. If $R$ is below 1, we would expect incidence to decline on average, but several cases could still occur before transmission ceases \citep{farringtonDistributionTimeExtinction1999}. Two additional factors can influence the probability a pathogen with $R$ above 1 will successfully establish within a population~\citep{lloyd-smithSuperspreadingEffectIndividual2005}. If there is a large degree of individual-level variation in transmission, the majority of infectious individuals will infect nobody, even though a minority will infect many; as a result, most newly introduced outbreaks will fail to take off. In addition, the likelihood of establishment increases if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\ % AK: Note 'primary case' means first in outbreak

Early in a novel outbreak, pharmaceutical options such as treatments and vaccines may not yet be available, leading to reliance on non-pharmaceutical interventions (NPIs) to reduce infection and disease. These interventions include individual-level targeted measures such as isolation and contact tracing, to population-level policies such as venue closures or restrictions on social mixing (Huag et al., 2020) [AK: probably a better reference out there, e.g. Sharma et al, Nature Comms]. The success of such interventions in curtailing transmission before it becomes sustained depends on several factors, including: the type of intervention, timing of introduction, and the intervention's effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with a short generation times (e.g. pandemic influenza), incidence of new infections can grow quickly, outpacing a response. Moreover, pathogens such as influenza, which have a large proportion of onward transmission prior to symptom onset and have asymptomatic (subclinical) cases, are less easily controlled via isolation of symptomatic cases and tracing and quarantining of their contacts~\citep{fraserFactorsThatMake2004}.

Several pre-COVID-19 studies concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and tracing their contacts, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. \\

@@

%The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\ [AK: These headline fatality risks are likely to be biased: https://kucharski.substack.com/p/how-fatal-is-h5n1-influenza]

In this study we evaluate the effectiveness of controlling outbreaks that exhibit influenza-like epidemiological characteristics using a mathematical model of isolation and contact tracing. In particular, we investigate the assumption that influenza will inevitably outpace ability to trace and isolate sufficiently effectively for control, and whether this assumption holds for all pathogen subtypes given emerging estimates of incubation period and serial interval \citep{Ward2024.12.11.24318702}. Finally, we quantify the speed and effectiveness of isolation for an effective response under different response delays across epidemic scenarios, and assess the impact of reducing the delay to isolation on outbreak containment alongside use of rapid testing.

@@

\subsection*{Transmission model}
We used a branching process transmission model to assess the potential effectiveness of isolation and contact tracing to control an outbreak of pandemic-potential influenza \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}). In the model, each infectious individual generates secondary cases from a negative binomial distribution with a given reproduction number $R$ and dispersion parameter ($k$). Onwards transmission from infectious individuals could be reduced in the following ways: 1) \textit{symptomatic individuals isolate effectively}, 2) \textit{contacts of symptomatic individuals are traced and isolate if they become symptomatic}, 3) \textit{contacts of symptomatic individuals are traced and effectively quarantined immediately}. Symptomatic individuals that have been traced (i.e. infectees) are isolated either 1) after the onset-to-isolation delay after symptom onset, or 2) when their infector is isolated. The isolation time of the infectee is whichever occurs first of these two events.

We include several mechanisms by which isolation and contact tracing can be imperfect at reducing onward disease transmission. The proportion of contacts identified by contact tracing can range between 0\% (i.e. no contacts are traced) to 100\% (i.e. every contact of every symptomatic case is traced). Because the model focuses on onwards transmission prevented, identifying 50\% of cases and preventing 100\% of their onwards transmission via contact tracing is equivalent to identifying 100\% of cases and preventing 50\% of their onwards transmission. Subsequent symptomatic cases that have not been contact traced are only isolated after the onset-to-isolation delay post symptom onset. Additionally, we assume infections have an independent probability of being asymptomatic or paucisymptomatic; these infections were not identified and therefore their contacts were not traced. The advantage of this approach is that it enables estimation of the overall population transmission effect of multiple interventions without requiring each step to be individually specified as a parameter.

In this study we evaluate the feasibility of controlling outbreaks of A/H5N1, A/H1N1, and A/H7N9-like infections.\cite{nishiuraEstimationIncubationPeriod2011} report the best fitting Weibull shape and scale parameters for H1N1 to be 1.77 and 1.86, respectively (Figure \ref{fig:delay-distributions}A). \cite{cowlingComparativeEpidemiologyHuman2013} report the incubation period for H5N1 and H7N9, both Weibull distributed. For H5N1 the Weibull distribution has a mean of 3.3 days and standard deviation (SD) of 1.5 days, for H7N9 the Weibull distribution has a mean of 3.1 days and SD 1.4 days \citep{cowlingComparativeEpidemiologyHuman2013} (Figure \ref{fig:delay-distributions}A). Comparatively, the incubation period for H1N1 shorter than the other two subtypes (H5N1 and H7N9), which are similar, with H5N1 having a slightly longer median (Figure \ref{fig:delay-distributions}A).  We consider four reproduction number ($R$) values when simulating transmission: 1.1, 1.5, 2.5 and 3.5. These are based on similar values to previously estimated $R$ values for influenza outbreaks \citep{fergusonStrategiesMitigatingInfluenza2006}, including H1N1 \citep{fraserPandemicPotentialStrain2009, lesslerOutbreak2009Pandemic2009}, or they represent inflated values of current estimates of H5N1 and H7N9 subtypes to evaluate the effectiveness of isolation and contact tracing in the event that pathogen evolution.

We three presymptomatic transmission scenarios: <1\%, 15\%, 30\% of transmission occurring prior to symptom onset. % AK: Probably need a 50%?
We assumed that asymptomatic infections can occur and infect others, but are less common that symptomatic infectious indivdiuals and that the majority of cases will develop symptoms during their infectious period.  We ran three scenarios, one with zero asymptomatic illness, another with 10\% of cases being asymptomatic and the most extreme scenario where 30\% of cases never develop symptoms and thus are not isolated and secondary cases are missed by contact tracing.

@@

We define outbreak control as no new cases between 12 and 16 weeks after the start of an outbreak\cite{hellewellFeasibilityControllingCOVID192020}. The rationale for this criteria is that some outbreaks that do not take off and become large may still have a few cases for the first few weeks of an outbreak, therefore the window for control needs to be enough time after the seeding of an outbreak to be sure it is extinct. There is also a maximum number of cases in simulation model, that when reached is judged to have reached a large epidemic and not controllable within 12-16 weeks. We set this limit to 500 cases. To test scenarios where an outbreak starts with different number of initial infections, we simulate with either 5, 20 or 40 initial independent cases. This could represent importation of multiple cases in a single event, such as a flight with multiple positive cases. \\

The onset-to-isolation delay is parameterised with three scenarios: a fast response, based on the SARS in Hong Kong (Donnelly et al. 2003) and a slower response based on COVID-19 in Wuhan, China (Li et al. 2020) (Figure \ref{fig:delay-distributions}B). Both the fast and slow onset-to-isolation are parameterised as Weibull distributions, the SARS-like distribution has shape ($\lambda$) 1.65 and scale ($k$) 4.29 (mean = 3.83 days, SD = 2.38 days), the COVID-like distribution has shape = 2.31 and scale = 9.48 (mean = 8.40, SD = 3.87 days) (Figure \ref{fig:delay-distributions}B). We also include a third onset-to-isolation scenario based on the case where lateral flow rapid tests (LFTs) are provided to all individuals in the susceptible population and these people are freely able to self-test if they suspect they are symptomatic. We parameterise this as an exponential distribution, with rate ($\lambda$) 0.5, where most individuals will self-test very shortly after they become symptomatic (approximately 40\% of cases will self-test within the first day of sympmtom onset), with a few people waiting a 2 or more days before testing (approximately 3\% take longer than one week to take a self-test). We assume that LFTs are 100\% accurate so all symptomatic individuals will receive a positive test result.

@@

The most rapid scenario we simulated to emulate individuals self-testing with LFTs, has the same trend of increased chance of outbreak containment by reducing the time between symptom onset and isolation, as the slow and fast scenarios (Figure \ref{fig:prop-outbreak-control-R}). The control of low transmissible ($R \in {1.1, 1.5})$ influenza pathogens is almost always guranteed even in the absence of contact tracing and only isolating cases immediately if positive on an LFT (Figure \ref{fig:prop-outbreak-control-R}C). Assuming ubiquitous access to home testing and that 63\% of symptomatic cases test and isolate within the 24 hours symptom onset, 75\% ascertainment of contacts results in most outbreaks controlled for all influenza transmissibility scenarios ($R \leq 3.5$). The LFT isolation strategy is also the only onset-to-isolation delay which can achieve 100\% of outbreaks controlled when all contacts are traced (Figure \ref{fig:prop-outbreak-control-R}C). \\

@@

\section*{Supplementary Materials}

To assess the efficacy of isolation and contact tracing to control an outbreak of pandemic-potential influenza we used an individual-level transsmision model implemented in the R package \texttt{\{ringbp\}} \citep{hellewellRingbpSimulateEvaluate2025} (Figure \ref{fig:ringbp-model}). The epidemic simulation model uses a single-type branching process with a negative binomial offspring distribution to simulate the spread of a pathogen through a population (infinite population size, without depletion of susceptibility). Influenza subtypes have shown moderate transmission heterogeneity \citep{fraserPandemicPotentialStrain2009, heComparingCOVID191918192020, Ward2024.12.11.24318702}, but evidence is limited and the negative binomial allows flexibility between highly heterogeneous and homogeneous individual-level transmission. The mean of the negative binomial offspring distribution is the effective reproduction number ($R$), and this distribution allows the model to be parameterised to capture overdispersion of transmission, in other words, some infectors are superspreaders and infect a disproportionate number of individuals \citep{lloyd-smithSuperspreadingEffectIndividual2005, kucharskiEarlyDynamicsTransmission2020}. \\

% AK EDIT: Align with definitions in main methods

The NPI implemented in the transmission model is an isolation of symptomatic cases and contact tracing (Figure \ref{fig:ringbp-model}). When a case is symptomatic they are isolated after a delay. The timing of isolation is drawn from the onset-to-isolation distribution (see \nameref{epiparameters} section for parameterisation). Any infection times that are sampled to occur after the isolated of an infector are removed, because isolation is assumed completely effective at preventing further transmission, reducing the number of secondary cases if the speed of isolation is quicker than some or all of the secondary infection times. In other words, once a case is isolated at time $t_{isolated}$, the probability of that case infecting anyone at $t > t_{isolated}$ is zero. Another way to interpret the model is that $R$ is composed of secondary infections before isolation ($R_{pre}$) and secondary infections after an assumed isolation time ($R_{isolated}$). Symptomatic individuals that are traced are isolated either when 1) the onset-to-isolation delay after symptom onset, or 2) the infector is isolated. The isolation of the infectee is whatever occurs first of these two events.  \\
